---
title: "Introduction to reutils"
author: "Gerhard Schöfl"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to reutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
hook_output <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) {
    if (length(x) > lines) {
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(if (abs(lines[1]) > 1 | lines[1] < 0) more else NULL,
           x[lines],
           if (length(x) > lines[abs(length(lines))]) more else NULL
    )
  }
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
}) 
options(reutils.api.key = NULL)
options(reutils.rcurl.connecttimeout = 50)
library(reutils)
```

`reutils` is an R package for interfacing with NCBI databases such as PubMed,
Genbank, or GEO via the Entrez Programming Utilities
([EUtils](https://www.ncbi.nlm.nih.gov/books/NBK25501/)). It provides access to the
nine basic *eutils*: `einfo`, `esearch`, `esummary`, `epost`, `efetch`, `elink`,
`egquery`, `espell`, and `ecitmatch`.

Please check the relevant
[usage guidelines](https://www.ncbi.nlm.nih.gov/books/NBK25497/#chapter2.Usage_Guidelines_and_Requiremen)
when using these services. Note that Entrez server requests are subject to frequency limits.
Consider obtaining an NCBI API key if are a heavy user of E-utilities.

## Important functions ##

With nine E-Utilities, NCBI provides a programmatical interface to the Entrez query and database system for searching and retrieving requested data

Each of these tools corresponds to an `R` function in the reutils package described below.

#### `esearch` ####

`esearch`: search and retrieve a list of primary UIDs or the NCBI History
Server information (queryKey and webEnv). The objects returned by `esearch`
can be passed on directly to `epost`, `esummary`, `elink`, or `efetch`.


#### `efetch` ####

`efetch`: retrieve data records from NCBI in a specified retrieval type
and retrieval mode as given in this
[table](https://www.ncbi.nlm.nih.gov/books/NBK25499/table/chapter4.T._valid_values_of__retmode_and/?report=objectonly). Data are returned as XML or text documents.

#### `esummary` ####

`esummary`: retrieve Entrez database summaries (DocSums) from a list of primary UIDs (Provided as a character vector or as an `esearch` object)

#### `elink` ####

`elink`: retrieve a list of UIDs (and relevancy scores) from a target database
that are related to a set of UIDs provided by the user. The objects returned by
`elink` can be passed on directly to `epost`, `esummary`, or `efetch`.

#### `einfo` ####

`einfo`: provide field names, term counts, last update, and available updates
for each database.

#### `epost` ####

`epost`: upload primary UIDs to the users's Web Environment on the Entrez
history server for subsequent use with `esummary`, `elink`, or `efetch`.


## Examples

### `esearch`: Searching the Entrez databases ###

Let's search PubMed for articles with Chlamydia psittaci in the title that have
been published in 2020 and retrieve a list of PubMed IDs (PMIDs).

```{r, eval=TRUE, echo=TRUE}
pmid <- esearch("Chlamydia psittaci[titl] and 2020[pdat]", "pubmed")
pmid
```

Alternatively we can collect the PMIDs on the history server.

```{r, eval=TRUE, echo=TRUE}
pmid2 <- esearch("Chlamydia psittaci[titl] and 2020[pdat]", "pubmed", usehistory = TRUE)
pmid2
```

We can also use `esearch` to search GenBank. Here we do a search for polymorphic
membrane proteins (PMPs) in Chlamydiaceae.

```{r, eval=TRUE, echo=TRUE, output.lines=20}
cpaf <- esearch("Chlamydiaceae[orgn] and PMP[gene]", "nucleotide")
cpaf
```

Some accessors for `esearch` objects

```{r, eval=TRUE, echo=TRUE}
getUrl(cpaf)
```


```{r, eval=TRUE, echo=TRUE}
getError(cpaf)
```

```{r, eval=TRUE, echo=TRUE}
database(cpaf)
```

Extract a vector of GIs:

```{r, eval=TRUE, echo=TRUE}
uid(cpaf)
```

Get query key and web environment:

```{r, eval=TRUE, echo=TRUE}
querykey(pmid2)

```

```{r, eval=TRUE, echo=TRUE}
webenv(pmid2)
```

Extract the content of an EUtil request as XML.

```{r, eval=TRUE, echo=TRUE, output.lines=20}
content(cpaf, "xml")
```

Or extract parts of the XML data using the reference class method `#xmlValue()`
and an XPath expression:

```{r, eval=TRUE, echo=TRUE}
cpaf$xmlValue("//Id")
```

### `esummary`: Retrieving summaries from primary IDs ###

`esummary` retrieves document summaries (*docsum*s) from a list of primary IDs.
Let's find out what the first entry for PMP is about:

```{r, eval=TRUE, echo=TRUE, output.lines=20}
esum <- esummary(cpaf[1])
esum
```

We can also parse *docsum*s into a `tibble`

```{r, eval=TRUE, echo=TRUE}
esum <- esummary(cpaf[1:4])
content(esum, "parsed")
```

### `efetch`: Downloading full records from Entrez ###

First we search the protein database for sequences of the 
**c**hlamydial **p**rotease **a**ctivity **f**actor, 
[CPAF](http://dx.doi.org/10.1016/j.tim.2009.07.007)

```{r, eval=TRUE, echo=TRUE}
cpaf <- esearch("Chlamydia[orgn] and CPAF", "protein")
cpaf
```

Let's fetch the FASTA record for the first protein. To do that, we have to
set `rettype = "fasta"` and `retmode = "text"`:

```{r, eval=TRUE, echo=TRUE}
cpaff <- efetch(cpaf[1], db = "protein", rettype = "fasta", retmode = "text")
cpaff
```

Now we can write the sequence to a fasta file by first extracting the data from
the `efetch` object using `content()`:

```{r, eval=TRUE, echo=TRUE}
write(content(cpaff), file = "~/cpaf.fna")
```

```{r, eval=TRUE, echo=TRUE, output.lines=20}
cpafx <- efetch(cpaf, db = "protein", rettype = "fasta", retmode = "xml")
cpafx
```

```{r, eval=TRUE, echo=TRUE, output.lines=20}
aa <- cpafx$xmlValue("//TSeq_sequence")
aa
defline <- cpafx$xmlValue("//TSeq_defline")
defline
```

### `einfo`: Information about the Entrez databases ###

You can use `einfo` to obtain a list of all database names accessible through
the Entrez utilities:

```{r, eval=TRUE, echo=TRUE}
einfo()
```

For each of these databases, we can use `einfo` again to obtain more information:

```{r, eval=TRUE, echo=TRUE}
einfo("taxonomy")
```
